The Form of Uncertainty Affects Selection for Social Learning

### 2023-04-04 ms.tex
fixed two Figure 5's issue
revision #2 updates

\begin{abstract} 
Social learning is a critical adaptation for dealing with different forms of variability. 
Uncertainty is a severe form of variability where the space of possible
decisions or probabilities of associated outcomes are unknown.
We identified four theoretically-important sources of uncertainty: temporal
environmental variability, payoff ambiguity, selection-set size, and effective lifespan. When these combine,
it is nearly impossible to fully learn about the environment. We develop an
evolutionary agent-based model to test how each form of uncertainty affects the evolution of social
learning. Agents perform one of several behaviors, modeled as a multi-armed bandit, to acquire
payoffs. \edit{All agents learn about behavioral payoffs individually through an adaptive
behavior-choice model that uses a softmax decision rule.} Use of vertical and oblique \edit{payoff-biased} social learning \edit{evolved} to serve as a scaffold \edit{for adaptive individual learning---they are not opposite strategies}. Different types of uncertainty had varying effects.
Temporal environmental variability suppressed social learning, whereas larger selection-set size
promoted social learning, even when the environment changed frequently. Payoff ambiguity and lifespan
interacted with other uncertainty parameters. \edit{This study begins to explain how social learning can predominate despite highly variable real-world environments when effective individual learning helps individuals recover from learning outdated social information.}
%, meaning social and individual learning complement one another
  % the importance of clearly operationalizing uncertainty for understanding its role in cultural evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
% \footnote{This document contains
% \total{citnum} references.}  
\end{abstract}

@@

To address these concerns we developed an evolutionary agent-based model that
simultaneously operationalizes several forms of uncertainty and endows agents
with a relatively powerful mechanism for individual learning. We model four kinds of
uncertainty that are common in cultural evolution models and the related empirical
literature:  (1) temporal environmental variability; (2) selection-set size (the number of possible behaviors); (3) payoff ambiguity
(the difference in the expected rewards for different behaviors); and (4) the effective lifespan of agents (the number of
opportunities for individual learning), see next section for more details. In our model, social learners learn from the previous
generation and asocial learners do not. All agents engage in individual learning
within their lifespans. The model also assumes \edit{an adaptive},
empirically-motivated learning mechanism, the softmax \edit{decision rule}, which allows learners to make full use of the available data acquired by both individual and social learning \citep{SuttonBartoBook}. Our model design thereby affords us a richer and possibly more accurate assessment of trade-offs between social and asocial learning. We use this integrated model to ask which kinds of uncertainty are likely to favor social learning and to clarify the logic underlying the results.
In the remainder of this introduction we briefly review previous research on both the forms of uncertainty and the learning mechanism used in our model.

@@

\paragraph{Payoff ambiguity.} Most models of social learning necessarily differentiate between the payoffs for adopting optimal vs. non-optimal behaviors
\citep{BoydRicherson1985,Rogers1988,Enquist2007,Rendell2010,aoki2014evolution}. The size of the difference between these payoffs is usually taken to influence the strength of selection on learning strategies.
However, the ability to discern payoff differences between behaviors is also a source of uncertainty. In reality, payoffs for particular behaviors are not always consistent. A behavior may yield a large payoff sometimes and a small
payoff other times \citep{McElreath2005}. This means that signals about the
relationship between behavior and payoff are often noisy, and differentiating
between behavioral options is in part a problem of signal detection. When the
difference in expected payoffs between optimal and non-optimal behaviors is very
large, this noise matters little, as the signal is still very clear. However, when
the expected payoffs of different behaviors are similar relative to the size of
their variances, ambiguity arises about which behaviors really are superior.
Smaller differences between payoffs corresponds to larger ambiguity. Payoff
ambiguity has been manipulated in both theoretical~\citep{perreault2012bayesian}
and empirical~\citep{McElreath2005,Toyokawa2019} studies, both of which support the
claim that payoff ambiguity increases the reliance on social information.
Importantly, payoff ambiguity %differs from selection-set size in that it not only 
affects both uncertainty \emph{and} the strength of selection, with smaller payoff differences leading to greater uncertainty for a learner and weaker evolutionary selection favoring optimal strategies. 
\edit{It is an outstanding problem to understand how payoff variance and payoff
distributions generally affect evolved behavioral strategies~\citep{Haaland2019}.
For the sake of theoretical clarity for our goal of understanding the
\emph{interaction} of various sources of uncertainty, we only operationalize payoff
ambiguity as the difference between expected payoffs, even though empirical studies
have sometimes included both forms of payoff uncertainty~\citep{McElreath2005,Toyokawa2019}, and a more complete model would include differences in mean and variance of payoffs.}

@@

Empirically, the number of learning opportunities can be manipulated in the lab and in the real world will tend to correlate with an individual's relative age. In multi-round studies of information use in novel tasks, US participants' use of social
information declined precipitously across rounds~\citep{McElreath2005}, suggesting
they were more likely to use social information when they were most uncertain about the task early on. In a more naturalistic context, \citet{Aplin2017} found that younger great tits more readily used social information compared to older individuals, possibly because they had accrued less information via individual learning, and possibly because younger individuals have the most to gain (because of higher reproductive value) by switching to superior behavioral options. Cross-cultural studies have highlighted the importance of childhood as a phase of heavy social learning in humans \citep{Reyes2016}. Young children are more likely to
acquire their beliefs and simple skills from their parents than are older children
or adults, which is at least partly due to the differential knowledge accrual
between young children and the older adults to whom they direct the most trust and
attention \citep{kline2013teaching}. While these phenomena map imperfectly onto effective lifespan in our model, given social learning is only allowed intergenerationally, \edit{the relative number of individual to social learning opportunities, i.e., this model's version of }effective lifespan, do vary across tasks, individuals, and species, yet most models assume only one learning opportunity per generation \citep{BoydRicherson1985, Feldman1996,Henrich1998, perreault2012bayesian}. Some models do allow several cultural generations within one genetic generation \citep{Enquist2007,Rendell2010,lindstrom2016co}, but little formal theory explicitly examines the role of learning opportunities in the evolution of social learning.

@@

Many models of social learning simplify by using minimally-cognitive agents. For example, a common modeling strategy compares the payoffs of agents with different pure learning strategies (e.g., those who only engage in social learning versus those who only engage in individual learning) while revealing nothing about the cognition underlying individuals' decisions~\citep{BoydRicherson1985, Rogers1988, 
aoki2005}. This sort of behavioral gambit is common in evolutionary modeling, but has also been criticized as `blackboxing' key cognitive processes that are important to cultural evolution \citep{Heyes2016,Kendal2018}.  
Some more complicated learning strategies that make use of both socially and
individually learned information have been studied ~\citep{Enquist2007,
perreault2012bayesian}, including those that integrate cognitively plausible
mechanisms such as reinforcement learning \citep{lindstrom2016co}, though the latter
approach remains the exception rather than the norm. In order to understand how
social learning evolves as an adaptation to uncertainty, we endowed agents with an
\edit{adaptive} cognition based on a softmax decision rule capable of integrating information from various sources to maximize payoffs~\citep{Gershman2019}, described in more detail below.

@@

The structure of the environment is specified by several parameters that remain fixed throughout the course of a given simulation (Table~\ref{tab:uncertaintyParameters}).  The environment affords $B$ behaviors, where $B$ is the \emph{selection-set size}. Uncertainty increases with $B$. Each behavior is indexed by $b$ and yields a payoff of 1 with probability $\pi_b$ and a payoff of 0 otherwise. As such, $\pi_b$ is equivalently the expected payoff of behavior $b$. At the beginning of each generation, one behavior is optimal and yields expected payoff $\pihigh$; the other $B-1$ behaviors yield a lower payoff, $\pilow$. The \emph{payoff ambiguity} is defined by $\pihigh - \pilow$. We fix $\pihigh = 0.9$ to be constant across all simulations, so payoff ambiguity is varied by changing the value of $\pilow$. In other words, payoff ambiguity, and therefore uncertainty, increases with $\pilow$.  Agents have $L$ time steps to \edit{act and learn asocially} per generation (i.e., the \emph{effective lifespan}). Uncertainty decreases with $L$. At the start of each generation a new optimal behavior is selected at random from the set of all possible behaviors with probability $u$ (the \emph{environmental variability}), otherwise the optimal behavior remains unchanged from the previous generation. Uncertainty increases with $u$.

@@

Agent $i$ using behavior $b$ receives a payoff of 1 with probability $\pihigh$
if $b$ is the optimal behavior, or with probability $\pilow$ otherwise. 
%Agents probabilistically recieve a payoff of 0 or 1 depending on whether the bandit paid off. Behavior payoffs are drawn from a Bernoulli distribution with mean $\pi_b$, so $\pi_b$ is also the mean payoff from behavior $b$.  For concreteness we denote this probabilistic payoff $\bandit$. 
After performing behavior $b$, agent $i$ updates its
corresponding \emph{behavior count} for $b$ by 1, $c_{ib}' \leftarrow c_{ib} + 1$.
Then the observed average payoffs known by agent $i$ for behavior $b$ are updated
via \edit{moving arithmetic averaging},
\begin{equation}
  \ledger' \leftarrow \ledger + \frac{\bandit - \ledger}{c_{ib}'},
\end{equation}
\noindent
where $\bandit$ is the actual payoff received by agent $i$ for behavior $b$ on a
given %(implicit) 
time step. \edit{This form allows the running mean payoffs to be calculated for each
behavior, weighting each observation equally, without knowing in advance how many
times a behavior will be performed. While more or less sophisticated averaging may
better represent specific species' abilities in different contexts, this form
provides a basic observational learning mechanism that may be further specified as
appropriate for different cases.}

\paragraph{Intergenerational dynamics.} Between generations, agents first reproduce
via asexual haploid reproduction, which determines the transmission of the social
learning trait. Social learners then learn from the previous generation, after which
all agents from the previous generation die off. % (Figure~\ref{fig:schematic}C). 
More
specifically, this all happens as follows.  $N$ reproducers are selected with
replacement over $N$ independent draws, biased by performance: 
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce in one draw}) =
\frac{\pi_i}{\sum_{i=1}^N \pi_i}.  
\end{equation} 
\noindent 
A child inherits its parent's social-learning trait $s_i$ without mutation.  A social learner child with $s_i = 1$ learns from a teacher from its parent's generation, including possibly their parent (i.e., both vertical and oblique learning are possible).  A child chooses its teacher by first randomly selecting $N_T$ prospective teachers from the population, then selecting the one in this set with the greatest payoff net, with ties broken randomly. By first subsetting $N_T$ prospective teachers we represent the fact that access to the entire population is not generally guaranteed, though this type of algorithm yields qualitatively similar results to ones that do not first produce subsamples \citep{smaldino2019open} and we show in the supplement that our results are robust to different $N_T$
(Figure~\ref{fig:nteachersSensitivity}). Social learner $i$ adopts their chosen 
teacher $j$'s observed average payoffs for each behavior, $b$, i.e., 
$\ledger \leftarrow \bar\pi_{jb}$; 
and agent $i$'s count of behavioral observations, $c_{ib}$, is set to 1 if teacher $j$ had at least
one observation of behavior $b$, and $c_{ib}$ is set to 0 otherwise. This way, expected payoffs
remain within $[0, 1]$, but social learners are flexible to adjust to environmental change. Setting
$c_{ib}$ to 1 reflects the fact that the social learner treats the teacher's information as a single
observation with the associated uncertainty. \edit{If learners have more confidence
in the teacher's information, or there are biological or cultural factors that more
strongly weight teacher information, it may be more justifiable to set the learner's $c_{ib}$ to its teacher's $c_{jb}$}. Asocial learners ($s_i = 0$) are initialized with
$c_{ib} = 0$ and $\ledger = 0$ for each behavior, $b$. % (Figure~\ref{fig:schematic}C). 
The newly initialized population then repeats the within-generation dynamics.

@@

\begin{figure}
  \caption{\edit{Model summary}. Agents are randomly initialized as social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation).
Otherwise it repeats another generation and evolution continues.} 
% \cm{What does (3) denote at end of Fig 1 caption? Does having only green agents in Fig1B give the impression that only social learners are learning within lifetime? Could the agent be half green/purple? Is it weird that in Fig1C a teacher with relatively low payoffs is being selected? Could be fixed by having the dotted line go from rich purple women's head to green baby's. }
  \label{fig:schematic}
  \centering
    \includegraphics[width=1.05\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied uncertainty values %(Table~\ref{tab:modelParameters}) 
and observed how frequently social learning evolved across 1000 trial simulations for each combination of parameters tested. %We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L \in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  
\edit{We observed three outcome measures: first, the social learning (SL) fixation frequency, denoted $\meansl$, is how frequently social learning evolved to fixation across all trials for a given parameter setting. To support our analysis about why we observed the patterns in social learning fixation frequency that we did, we also measured the mean number of generations to fixation, which measures the strength of selection.} Finally we measured the average net payoffs at the end of the simulation across agents and trials, normalized by lifespan. We analyze
outcomes by plotting $\meansl$ on the y-axis and environmental variability on
x-axis, since we theoretically expect that $\meansl$ will always decrease monotonically from 1 to 0 as $u$ increases. %Note that the hypothesis-testing concept of significance is meaningless here because we could make any small outcome difference ``significant'' by running more simulation trials.  We instead study the patterns of outcome variables over systematically varied uncertainty factor values.

@@

Our main results are illustrated by Figure~\ref{fig:mainResults}, which shows the
proportion of simulation runs that fixated at 100\% social learners as a function of
each of our four uncertainty measures. First, we found that reliance on social
learning monotonically decreases as temporal environmental variability, $u$,
increases \edit{(Figure~\ref{fig:mainResults})}. 
When the environment is more likely to change between generations,
information learned from the previous generation is less likely to be of value,
decreasing selection for social learning.  
However, the exact nature of the relationship between social
learning and temporal environmental variability was moderated by the other three uncertainty parameters.

\edit{Across all parameter settings, we observed that there exists some threshold value of environmental
  variability beyond which the evolution of social learning is exceedingly rare.  We call this the
  \emph{social learning ceiling}, denoted $u_\text{c}$ (Figure~\ref{fig:slCeiling}), which is expressed formally as 
\begin{equation}
  u_\text{c} = \min~\{ u~|~\meansl < 0.01 \}.
  \label{eq:slCeiling}
\end{equation}
\noindent
In other words, $u_\text{c}$ is the smallest value of $u$ for which social learning fixated
less than one in one hundred times. We will use this auxiliary measure to analyze the effect of
the other uncertainty parameters and to highlight that social learning frequently 
evolves despite high levels of environmental variability.}

The number of behavioral options, i.e., larger $B$, favors the evolution of social learning. Increasing the selection-set size
expands the range of $u$ over which social learning is favored for two reasons. First, when the environment remains constant between
generations, the value of social learning increases with more behavioral options, because it effectively increases the number of
observations that one can learn from by comparing the payoffs of several models. In other words, relying \emph{only} on one's
individual trial-and-error learning is less likely to yield useful information when the search space is
large. Second, when the environment changes, bias against the new optimal is weaker with more behavioral
options. For example, with only two behavioral choices ($B = 2$), a common assumption in many models, social
learning is particularly detrimental when the environment changes, since more of the population will have
arrived at the optimal behavior in the previous generation and socially learning from them will be biased
against the correct choice. These two mechanisms make it particularly profitable to engage in social
learning when the selection-set size is larger---for large values of $B$ social learning can even be favored
when the environment is more likely to change than not (i.e.,
$u>0.5$\edit{---see Figure~\ref{fig:slCeiling}}).

>>>>

\begin{abstract} 
Social learning is a critical adaptation for dealing with different forms of variability. 
Uncertainty is a severe form of variability where the space of possible
decisions or probabilities of associated outcomes are unknown.
We identified four theoretically-important sources of uncertainty: temporal
environmental variability, payoff ambiguity, selection-set size, and effective lifespan. When these combine,
it is nearly impossible to fully learn about the environment. We develop an
evolutionary agent-based model to test how each form of uncertainty affects the evolution of social
learning. Agents perform one of several behaviors, modeled as a multi-armed bandit, to acquire
payoffs. All agents learn about behavioral payoffs individually through an adaptive
behavior-choice model that uses a softmax decision rule. Use of vertical and oblique payoff-biased social learning evolved to serve as a scaffold for adaptive individual learning---they are not opposite strategies. Different types of uncertainty had varying effects.
Temporal environmental variability suppressed social learning, whereas larger selection-set size
promoted social learning, even when the environment changed frequently. Payoff ambiguity and lifespan
interacted with other uncertainty parameters. This study begins to explain how social learning can predominate despite highly variable real-world environments when effective individual learning helps individuals recover from learning outdated social information.
%, meaning social and individual learning complement one another
  % the importance of clearly operationalizing uncertainty for understanding its role in cultural evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
% \footnote{This document contains
% \total{citnum} references.}  
\end{abstract}

@@

To address these concerns we developed an evolutionary agent-based model that
simultaneously operationalizes several forms of uncertainty and endows agents
with a relatively powerful mechanism for individual learning. We model four kinds of
uncertainty that are common in cultural evolution models and the related empirical
literature:  (1) temporal environmental variability; (2) selection-set size (the number of possible behaviors); (3) payoff ambiguity
(the difference in the expected rewards for different behaviors); and (4) the effective lifespan of agents (the number of
opportunities for individual learning), see next section for more details. In our model, social learners learn from the previous
generation and asocial learners do not. All agents engage in individual learning
within their lifespans. The model also assumes an adaptive,
empirically-motivated learning mechanism, the softmax decision rule, which allows learners to make full use of the available data acquired by both individual and social learning \citep{SuttonBartoBook}. Our model design thereby affords us a richer and possibly more accurate assessment of trade-offs between social and asocial learning. We use this integrated model to ask which kinds of uncertainty are likely to favor social learning and to clarify the logic underlying the results.
In the remainder of this introduction we briefly review previous research on both the forms of uncertainty and the learning mechanism used in our model.

@@

\paragraph{Payoff ambiguity.} Most models of social learning necessarily differentiate between the payoffs for adopting optimal vs. non-optimal behaviors
\citep{BoydRicherson1985,Rogers1988,Enquist2007,Rendell2010,aoki2014evolution}. The size of the difference between these payoffs is usually taken to influence the strength of selection on learning strategies.
However, the ability to discern payoff differences between behaviors is also a source of uncertainty. In reality, payoffs for particular behaviors are not always consistent. A behavior may yield a large payoff sometimes and a small
payoff other times \citep{McElreath2005}. This means that signals about the
relationship between behavior and payoff are often noisy, and differentiating
between behavioral options is in part a problem of signal detection. When the
difference in expected payoffs between optimal and non-optimal behaviors is very
large, this noise matters little, as the signal is still very clear. However, when
the expected payoffs of different behaviors are similar relative to the size of
their variances, ambiguity arises about which behaviors really are superior.
Smaller differences between payoffs corresponds to larger ambiguity. Payoff
ambiguity has been manipulated in both theoretical~\citep{perreault2012bayesian}
and empirical~\citep{McElreath2005,Toyokawa2019} studies, both of which support the
claim that payoff ambiguity increases the reliance on social information.
Importantly, payoff ambiguity %differs from selection-set size in that it not only 
affects both uncertainty \emph{and} the strength of selection, with smaller payoff differences leading to greater uncertainty for a learner and weaker evolutionary selection favoring optimal strategies. 
It is an outstanding problem to understand how payoff variance and payoff
distributions generally affect evolved behavioral strategies~\citep{Haaland2019}.
For the sake of theoretical clarity for our goal of understanding the
\emph{interaction} of various sources of uncertainty, we only operationalize payoff
ambiguity as the difference between expected payoffs, even though empirical studies
have sometimes included both forms of payoff uncertainty~\citep{McElreath2005,Toyokawa2019}, and a more complete model would include differences in mean and variance of payoffs.

@@

Empirically, the number of learning opportunities can be manipulated in the lab and in the real world will tend to correlate with an individual's relative age. In multi-round studies of information use in novel tasks, US participants' use of social
information declined precipitously across rounds~\citep{McElreath2005}, suggesting
they were more likely to use social information when they were most uncertain about the task early on. In a more naturalistic context, \citet{Aplin2017} found that younger great tits more readily used social information compared to older individuals, possibly because they had accrued less information via individual learning, and possibly because younger individuals have the most to gain (because of higher reproductive value) by switching to superior behavioral options. Cross-cultural studies have highlighted the importance of childhood as a phase of heavy social learning in humans \citep{Reyes2016}. Young children are more likely to
acquire their beliefs and simple skills from their parents than are older children
or adults, which is at least partly due to the differential knowledge accrual
between young children and the older adults to whom they direct the most trust and
attention \citep{kline2013teaching}. While these phenomena map imperfectly onto effective lifespan in our model, given social learning is only allowed intergenerationally, the relative number of individual to social learning opportunities, i.e., this model's version of effective lifespan, do vary across tasks, individuals, and species, yet most models assume only one learning opportunity per generation \citep{BoydRicherson1985, Feldman1996,Henrich1998, perreault2012bayesian}. Some models do allow several cultural generations within one genetic generation \citep{Enquist2007,Rendell2010,lindstrom2016co}, but little formal theory explicitly examines the role of learning opportunities in the evolution of social learning.

@@

Many models of social learning simplify by using minimally-cognitive agents. For example, a common modeling strategy compares the payoffs of agents with different pure learning strategies (e.g., those who only engage in social learning versus those who only engage in individual learning) while revealing nothing about the cognition underlying individuals' decisions~\citep{BoydRicherson1985, Rogers1988, 
aoki2005}. This sort of behavioral gambit is common in evolutionary modeling, but has also been criticized as `blackboxing' key cognitive processes that are important to cultural evolution \citep{Heyes2016,Kendal2018}.  
Some more complicated learning strategies that make use of both socially and
individually learned information have been studied ~\citep{Enquist2007,
perreault2012bayesian}, including those that integrate cognitively plausible
mechanisms such as reinforcement learning \citep{lindstrom2016co}, though the latter
approach remains the exception rather than the norm. In order to understand how
social learning evolves as an adaptation to uncertainty, we endowed agents with an
adaptive cognition based on a softmax decision rule capable of integrating information from various sources to maximize payoffs~\citep{Gershman2019}, described in more detail below.

@@

The structure of the environment is specified by several parameters that remain fixed throughout the course of a given simulation (Table~\ref{tab:uncertaintyParameters}).  The environment affords $B$ behaviors, where $B$ is the \emph{selection-set size}. Uncertainty increases with $B$. Each behavior is indexed by $b$ and yields a payoff of 1 with probability $\pi_b$ and a payoff of 0 otherwise. As such, $\pi_b$ is equivalently the expected payoff of behavior $b$. At the beginning of each generation, one behavior is optimal and yields expected payoff $\pihigh$; the other $B-1$ behaviors yield a lower payoff, $\pilow$. The \emph{payoff ambiguity} is defined by $\pihigh - \pilow$. We fix $\pihigh = 0.9$ to be constant across all simulations, so payoff ambiguity is varied by changing the value of $\pilow$. In other words, payoff ambiguity, and therefore uncertainty, increases with $\pilow$.  Agents have $L$ time steps to act and learn asocially per generation (i.e., the \emph{effective lifespan}). Uncertainty decreases with $L$. At the start of each generation a new optimal behavior is selected at random from the set of all possible behaviors with probability $u$ (the \emph{environmental variability}), otherwise the optimal behavior remains unchanged from the previous generation. Uncertainty increases with $u$.

@@

Agent $i$ using behavior $b$ receives a payoff of 1 with probability $\pihigh$
if $b$ is the optimal behavior, or with probability $\pilow$ otherwise. 
%Agents probabilistically recieve a payoff of 0 or 1 depending on whether the bandit paid off. Behavior payoffs are drawn from a Bernoulli distribution with mean $\pi_b$, so $\pi_b$ is also the mean payoff from behavior $b$.  For concreteness we denote this probabilistic payoff $\bandit$. 
After performing behavior $b$, agent $i$ updates its
corresponding \emph{behavior count} for $b$ by 1, $c_{ib}' \leftarrow c_{ib} + 1$.
Then the observed average payoffs known by agent $i$ for behavior $b$ are updated
via moving arithmetic averaging,
\begin{equation}
  \ledger' \leftarrow \ledger + \frac{\bandit - \ledger}{c_{ib}'},
\end{equation}
\noindent
where $\bandit$ is the actual payoff received by agent $i$ for behavior $b$ on a
given %(implicit) 
time step. This form allows the running mean payoffs to be calculated for each
behavior, weighting each observation equally, without knowing in advance how many
times a behavior will be performed. While more or less sophisticated averaging may
better represent specific species' abilities in different contexts, this form
provides a basic observational learning mechanism that may be further specified as
appropriate for different cases.

\paragraph{Intergenerational dynamics.} Between generations, agents first reproduce
via asexual haploid reproduction, which determines the transmission of the social
learning trait. Social learners then learn from the previous generation, after which
all agents from the previous generation die off. % (Figure~\ref{fig:schematic}C). 
More
specifically, this all happens as follows.  $N$ reproducers are selected with
replacement over $N$ independent draws, biased by performance: 
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce in one draw}) =
\frac{\pi_i}{\sum_{i=1}^N \pi_i}.  
\end{equation} 
\noindent 
A child inherits its parent's social-learning trait $s_i$ without mutation.  A social learner child with $s_i = 1$ learns from a teacher from its parent's generation, including possibly their parent (i.e., both vertical and oblique learning are possible).  A child chooses its teacher by first randomly selecting $N_T$ prospective teachers from the population, then selecting the one in this set with the greatest payoff net, with ties broken randomly. By first subsetting $N_T$ prospective teachers we represent the fact that access to the entire population is not generally guaranteed, though this type of algorithm yields qualitatively similar results to ones that do not first produce subsamples \citep{smaldino2019open} and we show in the supplement that our results are robust to different $N_T$
(Figure~\ref{fig:nteachersSensitivity}).

Social learner $i$ adopts their chosen 
teacher $j$'s observed average payoffs for each behavior, $b$, i.e., 
$\ledger \leftarrow \bar\pi_{jb}$; 
and agent $i$'s count of behavioral observations, $c_{ib}$, is set to 1 if teacher $j$ had at least
one observation of behavior $b$, and $c_{ib}$ is set to 0 otherwise. This way, expected payoffs
remain within $[0, 1]$, but social learners are flexible to adjust to environmental change. Setting $c_{ib}$ to 1 reflects the fact that the social learner treats the teacher's information as a single observation with the associated uncertainty, \edit{and serves as a useful baseline for understanding the interaction of different sources of uncertainty that we focus on here. In the real world this could represent the fact that some amount of information is always lost in social learning, through instruction or otherwise.  If a learner's $c_{ib}$ is set instead to the teacher's $c_{jb}$, or some sizable fraction of it, learners will more slowly update their mean payoff values, making individual learning less adaptive, which could be desirable if there is good reason to have more confidence in a teacher's information, or there are other biological or cultural factors that make social learning more trustworthy.} Asocial learners ($s_i = 0$) are initialized with
$c_{ib} = 0$ and $\ledger = 0$ for each behavior, $b$. % (Figure~\ref{fig:schematic}C). 
The newly initialized population then repeats the within-generation dynamics.

@@

\begin{figure}
  \caption{Model summary. Agents are randomly initialized as social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation).
Otherwise it repeats another generation and evolution continues.} 
% \cm{What does (3) denote at end of Fig 1 caption? Does having only green agents in Fig1B give the impression that only social learners are learning within lifetime? Could the agent be half green/purple? Is it weird that in Fig1C a teacher with relatively low payoffs is being selected? Could be fixed by having the dotted line go from rich purple women's head to green baby's. }
  \label{fig:schematic}
  \centering
    \includegraphics[width=1.05\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied uncertainty values %(Table~\ref{tab:modelParameters}) 
and observed how frequently social learning evolved across 1000 trial simulations for each combination of parameters tested. %We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L \in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  
We observed three outcome measures: first, the social learning (SL) fixation frequency, denoted $\meansl$, is how frequently social learning evolved to fixation across all trials for a given parameter setting. To support our analysis about why we observed the patterns in social learning fixation frequency that we did, we also measured the mean number of generations to fixation, which measures the strength of selection. Finally we measured the average net payoffs at the end of the simulation across agents and trials, normalized by lifespan. We analyze
outcomes by plotting $\meansl$ on the y-axis and environmental variability on
x-axis, since we theoretically expect that $\meansl$ will always decrease monotonically from 1 to 0 as $u$ increases. %Note that the hypothesis-testing concept of significance is meaningless here because we could make any small outcome difference ``significant'' by running more simulation trials.  We instead study the patterns of outcome variables over systematically varied uncertainty factor values.

@@

Our main results are illustrated by Figure~\ref{fig:mainResults}, which shows the
proportion of simulation runs that fixated at 100\% social learners as a function of
each of our four uncertainty measures. First, we found that reliance on social
learning monotonically decreases as temporal environmental variability, $u$,
increases (Figure~\ref{fig:mainResults}). 
When the environment is more likely to change between generations,
information learned from the previous generation is less likely to be of value,
decreasing selection for social learning.  
However, the exact nature of the relationship between social
learning and temporal environmental variability was moderated by the other three uncertainty parameters.

Across all parameter settings, we observed that there exists some threshold value of environmental
variability beyond which the evolution of social learning is exceedingly rare.  We call this the
\emph{social learning ceiling}, denoted $u_\text{c}$ (Figure~\ref{fig:slCeiling}), which is expressed formally as 
\begin{equation}
  u_\text{c} = \min~\{ u~|~\meansl < 0.01 \}.
  \label{eq:slCeiling}
\end{equation}
\noindent
In other words, $u_\text{c}$ is the smallest value of $u$ for which social learning fixated
less than one in one hundred times. We will use this auxiliary measure to analyze the effect of
the other uncertainty parameters and to highlight that social learning frequently 
evolves despite high levels of environmental variability.

The number of behavioral options, i.e., larger $B$, favors the evolution of social learning. Increasing the selection-set size
expands the range of $u$ over which social learning is favored for two reasons. First, when the environment remains constant between
generations, the value of social learning increases with more behavioral options, because it effectively increases the number of
observations that one can learn from by comparing the payoffs of several models. In other words, relying \emph{only} on one's
individual trial-and-error learning is less likely to yield useful information when the search space is
large. Second, when the environment changes, bias against the new optimal is weaker with more behavioral
options. For example, with only two behavioral choices ($B = 2$), a common assumption in many models, social
learning is particularly detrimental when the environment changes, since more of the population will have
arrived at the optimal behavior in the previous generation and socially learning from them will be biased
against the correct choice. These two mechanisms make it particularly profitable to engage in social
learning when the selection-set size is larger---for large values of $B$ social learning can even be favored
when the environment is more likely to change than not (i.e.,
$u>0.5$---see Figure~\ref{fig:slCeiling}).

A shorter effective lifespan, $L$, also favors the evolution of social learning in
most cases. When effective lifespans are long and there are many opportunities to
gather information by individual learning, the value of social transmission is
diminished. Given enough opportunities, all agents will learn the optimal behavior
eventually. Thus, under most conditions, longer effective lifespans lead to a
narrower range of the parameters $u$ and $B$ under which social learning was
favored. It is noteworthy that in many cases, a large increase in the selection
regime for social learning is observed when $L=1$, and by definition no individual
learning can occur. The exceptions to this general pattern occur for high levels of
payoff ambiguity, $\pilow$ combined with low rates of environmental change, $u$
(left parts of plots on bottom row of Figure~\ref{fig:mainResults}). This reversal
is partially explained by weaker selection when there is little difference in the payoffs between
behaviors, and by the fact that the quality of social information is poorer when payoffs are ambiguous and
lifespans are short (more on these effects of payoff ambiguity below).  
When $L$ becomes roughly equal to or greater than $B$ there is little marginal effect of increasing $L$, indicating that more individual learning experiences are unlikely to significantly affect an agent's ability to identify the optimal behavior in their lifetime.
% \cm{is this right? I thought we should say something about the exceptions in this paragraph though we describe in more detail below, but not sure I have the right explanation}
